The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts

### 2020-05-04 write-up/covid-writeup.tex
covid update

\def\pr{\text{pr}}
\def\sgn{\text{sgn}}

@@

\begin{abstract}
Many statisticians, epidemiologists, economists and data scientists have registered serious reservations regarding the reported coronavirus case-counts. Limited testing capacity across many states has been widely identified as a key driver of suppressed coronavirus case-counts.  Calls for increased testing capacity are well-justified and increasingly frequent.  While expanded testing is laudable, the impact of selection bias does not disappear until .  Moreover, tests are imperfect.  The rate of false positives and false negatives interacts in complex ways with selection bias.  This note attempts to clarify this interaction.  We demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement-error. The discussion guides a sequence of suggestions in current case-count visualizations and reporting.    unadjusted prevalence estimates Measurement-error do not  misinterpret  that does not cover   testing, if not on  increases  election bias and measurement-error are often cited as key drivers of case count data.  It is known, for example, that false negatives deflate

However, most researchers forget about the importance of population sizes

In this note, we demonstrate through simple calculations how selection bias and theoretical
.

Comparing countries and states using those case-counts seem inappropriate when every nation/state have adopted different testing strategies and protocols. Here, we show the paradoxes that can persist  Estimating prevalence of COVID-19 based on these data is a hopeless exercise and several groups have recently argued for estimating the number of truly infected cases by using mortality rates.
% In a lucid article on FiveThirtyEight.com Nate Silver walks us through the problem by a simple numerical exercise in terms of the basic deterministic models of an epidemic using hypothetical scenarios involving R0 and the effective R.
In this note, we aim to (a)	we posit a conceptual mathematical framework to characterize both sampling bias and misclassification/imperfection of the test simultaneously on the estimation of the prevalence rate, (b)	review current testing strategies in some of the countries where we have testing data, and (c) provide guidelines for testing strategy/disease surveillance that may help track the pulse of the epidemic, to identify disease free areas and identify disease outbreaks.
% Also large-scale anti-body testing.
\end{abstract}

@@

Despite these necessary steps, the data landscape for understanding COVID-19 remains limited.  Public databases maintained by Johns Hopkins University (\url{https://bit.ly/2UqFSuA}) and the New York Times (\url{https://bit.ly/2vUHfrK}) provide incoming county-level information of confirmed cases and deaths.  Statisticians, epidemiologists, economists, and data scientists the world over have been using this granular data to

Due to limited testing capacity, local and state health departments have focused on testing only those from high-risk populations, resulting in low data quality due to bias in sampling from the overall population.

Despite these necessary steps, the data landscape for understanding COVID-19 remains limited.  Public databases maintained by Johns Hopkins University (\url{https://bit.ly/2UqFSuA}) and the New York Times (\url{https://bit.ly/2vUHfrK}) provide incoming county-level information of confirmed cases and deaths.

The WHO reports, however, only 103,945 total tests performed in the US as of March 19th.  Due to limited testing capacity, local and state health departments have focused on testing only those from high-risk populations, resulting in low data quality due to bias in sampling from the overall population.  The public health community requires auxiliary sources of information to improve national and local health policy decisions.  Certain survey efforts are underway, but may take time to yield fruit.

The goal of this paper is to express reservations at the use of case-counts as a proxy for prevalence and disease trajectory as well as its use as direct input into estimation of standard epidemiological models for inference and forecasting.  The two critical features are selection bias due to varied testing strategies and measurement error due to imperfect tests (i.e., false positives and negatives).  We demonstrate via theoretical analysis
More testing equals more data but does not imply more information.  The key issue is selection bias and measurement-error.

Given population sizes are quite heterogeneous,
the current data are of limited use to understand prevalence and even trajectory of the disease.

A critical question is to understand the mismatch between  there are alternative data streams that may be leveraged to understand the handling the COVID-19 pandemic in the US.

@@

Stanford study was Ionides.  They however, forgot about Critique #2: measurement

@@

This article discusses the relationship between three statistical concepts: selection bias, measurement-error, and the long-forgotten population size. We clarify mathematically why the situation is much more complex than it first appears.  Through simple mathematical argument, we demonstrate three important, yet often forgotten, tenants of

\begin{itemize}
	\item Unadjusted prevalence rates have a bias due to imperfect testing; bias is a complex function of measurement-error and
	\item Adjusted prevalence rates have an additional term in the error decomposition that highlights the interplay among measurement error, selection bias, and prevalence!
	\item All errors in these rates with population size so cross country comparisons are meaningless. Even if we scale by population size, the errors don't go away.  So comparing two countries of different sizes based on case-counts per million is difficult
	\item When testing increases are associated with increases in false positive/negative rates, the effective sample size may decrease.
	\item Comparison of daily trends in rate of positive tests is equally problematic.  When testing rates and selection protocols are held constant, increases are underestimated and
	\item Present a compartmental model that accounts for these time-varying factors.  Discuss identifiability (statistical issue) and how to proceed
	\item Review testing and building selection models (sensitivity analysis)
	\item Robustness (what directions are the models most resilient) and the notion of forecasts in bubbles
	\item Testing, disease surveillance and second-wave testing goals.
\end{itemize}

\section{}

The naive data analyst may think
\begin{center}
\emph{As we increase testing capacity, we will learn more about disease prevalence.}
\end{center}

Case-count is considered a proxy for disease prevalence and contagion spread.

Let $N$ denote the population size.  At a fixed moment in time, let $Y_j \in \{ 0,1\}$ for $j=1,\ldots, N$ denote COVID status.  As in survey sampling methodology, we treat these as fixed, unknown population quantities. For now, we ignore the dynamic nature of the viral outbreak; so either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$. A testing indicator, $I_j \in \{0,1\}$ is an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).

A primary question is the prevalence o
We are interested in the population average $\bar Y = \frac{1}{N} \sum_{j=1}^N Y_j$. Suppose we observe a sample of $n$ tests.  A natural candidate for prevalence is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$.

Prevalence is important in the long-run because the

Based on Meng (2019), we can express the error between $\bar y$ and the true proportion as
$$
\bar y_n - \bar Y =  \rho_{I, Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_Y
$$
Under random sampling $E [ \rho_{I,Y} ]$ = 0 and there is no bias.  Under selective testing,

Case-count
Figure~\ref{} presents testing per capita across reporting countries.  We see th at

Two countries with the same testing strategy (i.e., $\rho_{I,Y}$ equal) can yield wildly different estimates due to pop size.

For the current crisis, the analysis should be extended in two directions.  First, imperfect testing . Measurement-error.

@@

Tests are imperfect.  COVID-19 testing is no exception.  Data analysis must account for test inaccuracy to distinguish signal and noise.  Here we investigate the interplay between imperfect testing and selection bias.  When discussing testing inaccuracies, the standard assumption is measurement-error leads to parameter attenuation.  When paired with selection bias, however, the two sources of error become entangled, and the biases can become magnified, muted, or even switch signs.

First we require some additional notation.  Let $P_j$ be an indicator of measurement error, equal to $1$ when we mismeasure the outcome and $0$ when we observe the true outcome. We suppose this is a stochastic variable that satisfies $\pr(P_j = 1 \mid Y_j = 1, I_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0, I_j = 1) =: FP$ is the false-positive rate.  If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^\star = Y_j(1-P_j) + (1-Y_j) P_j$.  Suppose disease prevalence was estimated as the fraction who tested positive for COVID-19, i.e., $\bar y_n^\star = \frac{\sum_{i=1}^N I_j Y_j^\star}{\sum_{i=1}^N I_j}$.  We can again investigate the error compared to the true prevalence $\bar Y$ in statistical terms:
$$
\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left[ \rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} + \sqrt{\frac{f}{1-f}}  \left( FP - (FP+FN) \bar Y \right) \right] .
$$
where $Z = 1-2Y$. The first term represents the perfect testing regime, the second term represents the interaction between imperfect testing and selection bias, while the third time represents the bias due to imperfect testing.  From here on, we refer to $\rho_{I,Y}$ as the \emph{true data quality}, and $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement-error.  We show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, and so the second term adjusts the true data quality.

We start by considering the first two terms and assess whether the sign of the bias can reverse due to the interaction of measurement-error and selection bias.  To do this, we require the sampling rates differential.  Let $f_1 := P_J(I_J = 1 \mid Y_J = 1)$ and $f_0 := P_J(I_J = 1 \mid Y_J = 1)$ be the sampling rates.  Then if $\Delta = f_1 - f_0$ is the sampling rate differential, we have
$$
\rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} =
\rho_{I,Y} \times \sigma_Y \left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right].
$$
The final term in brackets is the \emph{measurement-error adjustment to data quality} which is a complex function of sampling rate differential, the odds ratio, and the ratio of measurement-error interaction with prevalence and sampling rates interaction with prevalence. Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ so the measurement-error adjustment either shrinks the data quality measure toward zero or reverses its sign.

While prior investigations have notes the interaction between measurement-error and selection bias (cites), the interaction with the sample size relative to the population,i.e., $f$, has largely been ignored.  The above statistical decomposition clarifies the importance of this quantity~$f$.  In particular, note that the statistical error also includes a bias term due to measurement-error and this term increases as the sampled fraction $f$ increases. Therefore, how the first two terms interact with the final term depends on the fraction of the population sampled.  This interaction is complex, but implies that whether the estimate $\bar y^\star_n$ is an overestimate or underestimate is a complicated question due to the relation amongst these three pieces.

The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples and, if sensitivity and specificity were known a priori, may suggest the alternative estimator $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling. We again wish to express the error $\tilde  y_n - \bar Y$ in statistical terms. In the appendix, we show that the error now can be expressed as
$$
\tilde y_n - \bar Y = \rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M}.
$$

The updated formula accounts for measurement-error to re

This results in an updates Z-score:
$$
Z_{n,N}
$$

Above we ignore an important consideration, namely that testing quality is likely a dynamic process.  In the US, states are rightly trying to increase testing capacity.  Commentators have requested expedited FDA approval for variations on the currently available tests to speed up production.  This is likely to increase testing inaccuracies, impacting both false negative and false positive rates.  Increased capacity may reduce

\begin{wrapfigure}{r}{0.5\textwidth}
\centering
\includegraphics[width = 0.4\textwidth]{../methods/figs/mem_heatmap_article.png}
\caption{Measurement-error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.15$ and $FN=0.10$. Color scaled so blue = $-6$, white = 0, and red = $6$.}
\end{wrapfigure}

For fixed prevalence,

The first term is the same as before but increased by $(1 + FP + FN)$ to account for the additional uncertainty due to measurement-error.  The second term is the interaction between selection bias and measurement error.  Interestingly we can show that the sign is reversed, i.e., $sgn(\rho_{I,PZ}) = 1 - sgn(\rho_{I,Y})$, leading to an absolute \emph{reduction} in the error.

@@

The effective sample size can then be bounded by
$$
n_{eff} = \frac{f}{1-f} \times E_{I} \left[ \rho_{I,Y}^2 D_M^2 \right]
$$

\subsection{Testing strategies and population size: a tango}

>>>>

\def\pr{\text{pr}}
\def\sgn{\text{sgn}}
\def\I{\bf I}

@@

\begin{abstract}
Many statisticians, epidemiologists, economists and data scientists have registered serious reservations regarding the reported coronavirus case-counts. Limited testing capacity across many states has been widely identified as a key driver of suppressed coronavirus case-counts.  Calls for increased testing capacity are well-justified and increasingly frequent.  While expanded testing is a laudable goal, selection bias will impact estimates of disease prevalence and the effective reproduction number until the entire population is sampled.  Moreover, tests are imperfect.  False positive/negative rates interact in complex ways with selection bias.  This note attempts to clarify this interaction.  Through simple calculations, we demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement-error. The discussion guides a series of suggestions to improve current case-count reporting.
\end{abstract}

@@

Despite these necessary steps, the data landscape for understanding COVID-19 remains limited.  Public databases maintained by Johns Hopkins University (\url{https://bit.ly/2UqFSuA}) and the New York Times (\url{https://bit.ly/2vUHfrK}) provide incoming county-level information of confirmed cases and deaths.  Statisticians, epidemiologists, economists, and data scientists the world over have been using this granular data to build model-based forecasts of future COVID-19 case-counts, deaths, and hospitalizations.  Variations on the SIR models have been used to draw inferences and build forecasts.  In most cases, parameter inference uses observed case-count and/or observed COVID-19 related deaths to infer latent trajectories of the pandemic (cites).

The goal of this paper is to express reservations at the use of case-counts as a proxy for prevalence and disease trajectory as well as its use as direct input into estimation of standard epidemiological models for inference and forecasting.  The reason is simple: current models ignore selection bias and measurement-error.

Selection bias enters due to differences in testing strategies across countries and states.  In the US, for example, limited testing capacity has caused local and state health departments to focus on testing only  high-risk populations.  Testing requires the individual to self-select into testing.  Demands for testing increases, while laudable, ignore the issues of self-selection and measurement-error. While increasing testing capacity increases data quantity, there is no guarantee for increased data quality.  We will show that aggressive pushes for ramped up testing capacity that are tied to decreases in data quality may have a deleterious effect in our ability to estimate quantities of interest such as prevalence and effective reproduction number.

In this paper, we will demonstrate the complex interaction between these two fundamental concepts.  Measurement-error is often associated with parameter attenuation; howeve,r we show that the combination with selection effects can cause bias in parameter estimates to change sign and increase/decrease in magnitude.  Moreover, the bias may depend on the sampling fraction, prevalence, and population size.  Without a complete understanding of these interations, we are destined to misinterpret case-counts and come to erroneous conclusions.   We will show, through simple calculations, how we can misguide ourselves and the need for more precision and care in presenting COVD-19 statistical models to the broader research and non-academic communities.

@@

Stanford study was Ionides.  They however, forgot about Critique \#2: measurement

@@

This article discusses the relationship between three statistical concepts: selection bias, measurement-error, and the long-forgotten population size. We clarify mathematically why the situation is much more complex than it first appears.  Through simple mathematical argument, we demonstrate five important, yet often forgotten, tenants of statistical science.  First we show that unadjusted prevalence rates are unsurprisingly biased when tests are imperfect. What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.  Next we show that data quality for adjusted prevalence rates incorporate an additional term in the error decomposition that highlights the interplay among measurement error, selection bias, and prevalence.  Third, we show that daily trends in rate of positive tests is equally problematic even if testing rates and selection protocols are held constant.  We show that the rate of change in COVID-19 observed case-counts are often an underestimate of the true rate. This implies use of observed rates will lead a data analyst to be overly optimistic in the early stages that the disease is not an issue, and also overly optimistic in the decline of the prevalence after the peak.   Fourth, we show that estimates of the reproducility rate are also impacted.

The above analyses are focused on the issues with using observed case-counts to understand the trajectory for a single country or state.  The media and organizations are also interested in comparisons of case-counts across countries. To account for population size, case-counts are scaled by the country population size.  We show that even after this adjustment, cross-country comparisons can be plagued by estimation error scaling by population size.  We end the paper with a discussion of the benefits of randomized testing, where auxiliary information is most relevant, and the distinction between data quality and decision making in the current pandemic.

\section{Analysis of case-count data}

We start with some simple notation.  Let $N$ denote the population size.  For state-level analysis $N$ is the state's total population, while for country-level analysis $N$ is the country's total population.  At a fixed moment in time, let $Y_j$ denote COVID-19 status for the $j$th individual in the population, $j=1,\ldots, N$. Here, like in survey methodology (cite), we treat COVID-19 status as a fixed but unknown population quantity of interest. For simplicity, we start by ignoring the dynamic nature of the viral outbreak as well as the fact that individuals can recover from the disease and assume either individual $j$ is COVID-19 positive and $Y_j=1$ or is COVID-19 negative and $Y_j=0$. We also let $I_j \in \{0,1\}$ be an indicator that the individual was selected for testing ($I = 1$) or not ($I=0$).

To start, we assume the primary questions of interest are estimates of the overall number of COVID-19 cases and/or disease prevalence. That is, we are interested in either the population total $Y = \sum_{j=1}^N Y_j$ or the population average $\bar Y = Y/N$. Suppose that $n$ tests are performed and we observe the values $y_1, \ldots, y_n \in \{0,1\}$.  Then a natural candidate for prevalence is the proportion of positive tests $\bar y = \frac{1}{n} \sum_{i=1}^n y_i$, and a natural candidate for overall cases is $N \times \bar y$.
Under simple random sampling (SRS) or any other epsem design, the above are unbiased estimators of the population-level quantities of interest.  Under SRS, the variance of the estimator can be expressed as $\frac{1}{N-1} \times \frac{1-f}{f} \times \sigma_Y^2$ where $f = n/N$ is the sampling fraction and $\sigma_Y^2 = \frac{1}{N} \sum_{i=1}^N (Y_i - \bar Y)^2 = \bar Y (1- \bar Y)$.

When the sampling mechanism is non-random, selection effects may cause bias in the above estimates. To better understand this issue, Meng (2019) provided the following intuitive and powerful statistical decomposition of the error between $\bar y$ and the true proportion $\bar Y$
$$
\bar y_n - \bar Y =  \rho_{I, Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_Y.
$$
The first term represents \emph{data quality}, the second \emph{data quantity}, and the third \emph{problem difficulty}. The term $\rho_{I,Y}$ is the empirical correlation between the population values~$\{ Y_j \}_{j=1}^N$ and the selection values $\{ I_j \}_{j=1}^N$.  Under simple random sampling, $E_{\I} [ \rho_{I,Y} ] = 0$, where the expectation is with respect to the selection mechanism~$\I$, so there is no bias.  The SRS variance formula above shows that $E_{\I} [ \rho_{I,Y}^2 ]  = 1/(N-1)$.  The key issue with selective testing is that $E_{\I} [ \rho_{I,Y} ] \neq 0$.  Meng identified this as the fundamental issue that can lead to paradoxes in the analysis of big data.

Here we highlight the core insights relevant to the COVID-19 crisis.  First, comparing the mean-squared error under selection mechanism $\I$ and SRS, we see that
$$
\frac{E_{\I} \left[ (\bar y_n - \bar Y)^2 \right]}{\sqrt{V_{SRS} (\bar Y)}} = (N-1) E_{\I} \left[ \rho_{I,Y}^2 \right].
$$
This points to a troubling and paradoxical situation: the error relative to SRS increases as a function of population size.  Meng terms this the ``Law of Large Populations'' (LLP).  The points to a critical issue in the current media practices in communication of case-count data: two countries with the same testing strategy (i.e., $\rho_{I,Y}$ equal) can yield wildly different estimates due to population size.  Large countries like the US may have similar true prevalence rates as smaller countries like the UK.  Even under similar testing strategies, the mean-squared error in the prevalence rate for the US will have substantially more variation.  For example, the MSE will be almost 5 times that in the US compared to the UK.  Therefore, strong opinions built from case-count records may be substantially distorted.

Second, there have been calls for increased testing.  While important, many conflate increased testing capacity with increased quality of testing.  We will discuss this point further in Section XX.  In the COVID-19 crisis, for example, New York has performed a total of $n = 872,481$ tests from April 20th to 30th.  In Figure XX, we present the trajectory of testing per day and positive cases per day.  The New York population is $N=8,399,000$; removing those individuals who have been already tested leaves a population of $7,XXX,XXX$.  This leads to a fraction sampled of $f = 0.1$.  A recent study attempted to estimate prevalence in New York.  After appropriate adjustments (see Appendix XX for details), a plausible range for prevalence seems to be between $2\%$ and $12\%$.  Suppose that individuals who are COVID-19 positive are $1.5$ times more likely to get tested than those individuals who are COVID-19 negative (a modest relative rate).  The question is ``What is the sample size from a SRS that would yield equivalent MSE in the estimated prevalence?'' Using the above MSE result, Meng showed the effective sample size is equal to $n_{eff} \leq \frac{1}{E_I [ \rho_{I,Y}^2 ]} \frac{f}{1-f}$ and that ranges from $208$ to $38$ across the range of plausible values of prevalence.  This is the sad state of affairs, saying that the effective sample size even in best case scenarios is not better than a small random sample from the population.

The remainder of this paper aims to build upon these fundamental insights by extending the decomposition in two directions: accounting for  measurement-error and the temporal nature of the pandemic.  We then discuss testing strategies, effective sampling methods, and data versus decision-making.

@@

Tests are imperfect.  COVID-19 testing is no exception. Here we investigate the interplay between imperfect testing and selection bias.  When discussing testing inaccuracies, the standard assumption is measurement-error leads to parameter attenuation.  When paired with selection bias, however, the two sources of error become entangled, and the biases can become magnified, muted, or even switch signs.

First we require some additional notation.  Let $P_j$ be an indicator of measurement error, equal to $1$ when we mismeasure the outcome and $0$ when we observe the true outcome. We suppose this is a stochastic variable that satisfies $\pr(P_j = 1 \mid Y_j = 1) =: FN$ is the false-negative rate and $\pr(P_j = 1 \mid Y_j = 0) =: FP$ is the false-positive rate.  If individual $j$ is selected (i.e., $I_j = 1$) then the observed outcome can be written as $Y_j^\star = Y_j(1-P_j) + (1-Y_j) P_j$.  Suppose disease prevalence was estimated as the fraction who tested positive for COVID-19, i.e., $\bar y_n^\star = \frac{\sum_{i=1}^N I_j Y_j^\star}{\sum_{i=1}^N I_j}$.  We can again investigate the error compared to the true prevalence $\bar Y$ in statistical terms:
$$
\bar y_n^\star - \bar Y = \sqrt{\frac{1-f}{f}} \left[ \rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} + \sqrt{\frac{f}{1-f}}  \left( FP - (FP+FN) \bar Y \right) \right] .
$$
where $Z = 1-2Y$. The first term represents the perfect testing regime, the second term represents the interaction between imperfect testing and selection bias, while the third time represents the bias due to imperfect testing.  From here on, we refer to $\rho_{I,Y}$ as the \emph{true data quality}, and $\rho_{I,PZ}$ as the \emph{observed data quality} that accounts for both selection bias and measurement-error.  We show the sign of $\rho_{I,PZ}$ is the opposite of the sign of $\rho_{I,Y}$, and so the second term adjusts the true data quality.

We start by considering the first two terms and assess whether the sign of the bias can reverse due to the interaction of measurement-error and selection bias.  To do this, we require the sampling rates differential.  Let $f_1 := \pr (I_J = 1 \mid Y_J = 1)$ and $f_0 := \pr(I_J = 1 \mid Y_J = 1)$ be the sampling rates.  Then $\Delta = f_1 - f_0$ is the sampling rate differential and we have
$$
\rho_{I,Y} \times \sigma_Y + \rho_{I,PZ} \times \sigma_{PZ} =
\rho_{I,Y} \times \sigma_Y \left[ 1 - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right].
$$
The final term in brackets is the \emph{measurement-error adjustment to data quality} which is a complex function of sampling rate differential, the odds ratio, and the ratio of measurement-error interaction with prevalence and sampling rates interaction with prevalence. Note $\sgn(\Delta) = \sgn(\rho_{I,Y})$ by equation~\ref{eq:binaryrho} so the measurement-error adjustment either shrinks the data quality measure toward zero or reverses its sign.

While prior investigations have noted the interaction between measurement-error and selection bias (cites), the interaction with the sample size relative to the population, i.e., $f$, has largely been ignored.  The above statistical decomposition clarifies the importance of this quantity~$f$.  In particular, note that the statistical error also includes a bias term due to measurement-error and this term increases as the sampled fraction $f$ increases. Therefore, how the first two terms interact with the final term depends on the fraction of the population sampled.  This interaction is complex, but implies that whether the estimate $\bar y^\star_n$ is an overestimate or underestimate is a complicated question due to the relation amongst these three pieces.

In the current setting, supposing the ratio of conditional selection rates $f_1/f_0$ is equal to 1.5.  The false negative rate ranges from 15\% to 40\%, while the false positive rate is low (around 1\%). Then under plausible prevalence rates that we discussed before, the relative MSE ranges from 1.4 to 0.12.  So that means in some cases we have a huge increase in MSE and in other settings we have a huge decrease in MSE.  What drives this is the false negative rate interaction with prevalence and sampling rates.  Therefore, whether we are better or worse off is a very difficult question to answer.

The attentive data analyst will recognize the estimator $\bar y_n$ is biased even for simple random samples and, if sensitivity and specificity were known a priori, may suggest the alternative estimator $\tilde y_n = \bar y_n + (1-\bar y_n) FP + FN \bar y_n$ which is unbiased under simple random sampling. We again wish to express the error $\tilde  y_n - \bar Y$ in statistical terms. In the appendix, we show that the error now can be expressed as
$$
\tilde y_n - \bar Y = \rho_{I,Y} \times \sqrt{\frac{1-f}{f}} \times \sigma_{Y}
\times \underbrace{\left[ 1 + FP + FN - \Delta \times \frac{\bar Y}{1-\bar Y} \times \frac{FP(1-\bar Y) + FN \cdot \bar Y}{f_0 (1-\bar Y) + f_1 \bar Y} \right]}_{D_M}.
$$
The first term is the same as before but increased by $(1 + FP + FN)$ to account for the additional uncertainty due to measurement-error.  The second term is the interaction between selection bias and measurement error.  Interestingly we can show that the sign is reversed, i.e., $sgn(\rho_{I,PZ}) = 1 - sgn(\rho_{I,Y})$, leading to an absolute \emph{reduction} in the error.

@@

\begin{wrapfigure}{r}{0.5\textwidth}
\centering
\includegraphics[width = 0.4\textwidth]{../methods/figs/mem_heatmap_article.png}
\caption{Measurement-error data quality adjustment: relative frequency $f_1/f_0$ (x-axis) against odds ratio (y-axis) for $FP=0.15$ and $FN=0.10$. Color scaled so blue = $-6$, white = 0, and red = $6$.}
\vspace{-0.7cm}
\end{wrapfigure}

Under the same numbers, $D_M$ is approximately 1.7.  The effective sample size can then be bounded by
$$
n_{eff} = \frac{f}{1-f} \times \frac{1}{E_{I} \left[ \rho_{I,Y}^2 D_M^2 \right]}.
$$
This lowers the ranges by a factor of almost $3$, leading to effective sample sizes from 69 to 13.

The above argument shows that the effective sample size is affected in a complex manner by the interaction between selection bias and measurement error.  Here, we discuss the impact of the interaction on the effective sample size when testing capacity is tied to measurement-error.  In the current pandemic, there have been well-justified calls for increases in testing capacity.  These increases, however, may come at the cost of increases in false-positive and false-negative rates.  In terms of effective sample size, if the data quality term remains constant, then testing increases can actually hurt effective sample size.  Consider the simplest case where $f$ increases from $0.05$ to $0.1$; suppose this required including tests such that the false negative rate jumps from $0.05$ to $0.15$.  Then we would think effective sample size should increase by a factor of $3.4$ but the actual increase is by a factor of only $1.8$!  Think about what this is saying, this is saying that sampling $15\%$ of the population rather than $5\%$ will only increase the effective sample size by a factor $1.8$ if the selection-bias persists and the false negative rate climbs to $15\%$.   It has been reported that false negative rates could increase more than that (cite NYT) and the false positive rate could go up slightly as well.  In a worst case scenario, if $FN \to 0.3$ and $FP \to 0.05$ then the factor is only 1.3.

\subsection{Regrettable rates}
\label{section:rates}

The data analyst, now frustrated with me but at least wary of estimating prevalence and total counts, pauses and thinks.  They return shortly thereafter, a bit agitated but persistent: \emph{Ok, perhaps total counts is a lost cause. Certainly, however, we can estimate the rate of growth.  All I want to know is when we hit the point at which the curve flattens and number of deaths decrease.  That can't be too hard, surely!}

Unfortunately, ratio estimators do not cancel errors as now both the numerator and the denominator are uncertain.  This means selection bias and measurement error can have paradoxical effects.  Here we consider the ratio estimators for the relative change in the prevalence rate.  We let $\bar Y_1$ and $\bar Y_2$ denotes the prevalence at time step $1$ and $2$ respectively (i.e., prevalence on two consecutive days).  Then the ratio estimator is given by $r = \bar y_1 / \bar y_2$.  We assume both numerator and denominator are the prevalence estimates adjusted for measurement-error. If the sample size at each time were equal (i.e., $n_1 = n_2 = n$) then this analysis would be equivalent to comparing the increase in observed case-counts (i.e., $\bar y_1/\bar y_2 = y_1/y_2$).  We wish to express the error $\frac{\bar y_2}{\bar y_1} - \frac{\bar Y_2}{\bar Y_1}$ in statistical terms.  Using a Taylor series approximation, we show that the error can be expressed approximately as
$$
\begin{aligned}
\frac{\bar Y_2}{\bar Y_1} \left[ \rho_{I_2, Y_2} D_{M_2} \sqrt{\frac{1-f_2}{f_2}} CV(Y_2)  - \rho_{I_1,Y_1} D_{M_1} \sqrt{\frac{1-f_1}{f_1}} CV (Y_1) \left[ \rho_{I_2, Y_2} D_{M_2} \sqrt{\frac{1-f_2}{f_2}} CV(Y_2) + 1 \right] \right]
\end{aligned}
$$
where $\rho_{I_j, Y_j}$ is the data quality, $f_j$ is the sampling fraction, $D_{M_j}$ is the measurement-error adjustment, and $CV(Y_j) = \sigma_{Y_j}/\bar Y_j$ is the coefficient of variation at time step $j$.  We see that the magnitude of the error depends on the population-level ratio $\bar Y_2 / \bar Y_1$ so a large decrease in the prevalence rate will have a relatively small error as compared to a large increase in the prevalence rate. There are three terms in the brackets.  The two terms $\rho_{I_2, Y_2} D_{M_2} \sqrt{\frac{1-f_2}{f_2}} CV(Y_2) - \rho_{I_1, Y_1} D_{M_1} \sqrt{\frac{1-f_1}{f_1}} CV(Y_1)$  represent the ``cancellation'' that the naive analyst is hoping will occur.  We can see that when the data quality, sampling fraction, and measurement-error adjustments are help constant this different simplifies greatly to $\rho_{I,Y} D_M \sqrt{\frac{1-f}{f}} ( \bar Y_1 - \bar Y_2)$. The final component $\bar Y_1 - \bar Y_2$ represents impact of the prevalence trajectory on the statistical error.  When prevalence is increasing the component is negative, and when prevalence is decreasing the component is positive. At the peak this term is zero, representing when the naive analysts intuition is closest to correct.  Regardless, we always observe the final term which is a cross term that depends on the statistical error in both stages.

Figure XX shows a trajectory of bias under several prevalence and data quality scenarios.  We see that in each case, the rate of change in prevalence is underestimated at most times, except for large decreases which are overestimated.  This points to a very sad realization.  At the beginning of the pandemic, we are underestimating the rate of change based on observed case counts.  Not only that, but the error is at its largest when the increase is large. This leads us to not take the virus seriously early on based on the observed data! Moreover, after the peak, we see that the bias is also negative other than for large decreases in the rate of change in prevalence.  This implies our estimates based on observed data trick us into thinking that the virus is ``going away'' faster than it actually is!

A final question is whether an observed drop in prevalence can be interpreted as a ``real'' decrease in prevalence.  To this end, one would typically construct a Z-score and perform a

Moreover, the question is ``Can we trust the observed data to let us know if we have reached the peak?''

Based on this analysis, seeing a falling rate for XX time steps in a row would be needed to feel confidence that the true prevalence rate is on the decline.

\subsubsection{Estimation of effective reproduction number}
Many well-respected epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis (cites).  We show here how these estimates are also impacted by selection bias and measurement error.  For simplicity, we again assume the estimates are adjusted for known false positive and false negative rates. Luis and Ribeiro (2008) show that under a Poisson likelihood, a simple relation between the trajectory of \emph{true new cases} and the effective reproduction number can be derived (see Appendix XX for discussion).  In particular, under an SIR model, the number of case counts is Poisson distributed with rate $K_{t-1} \exp \left( \frac{1}{\gamma} (R_t - 1) \right)$ where $\gamma$ is the serial interval, which is approximately $7$ days for COVID-19 (Sanche, 2020).  Using this, a method of moments estimator of the effective reproduction number at time $t$ can be given by
$$
R_t = 1 + \frac{1}{\gamma} \log \left( \frac{K_t}{K_{t-1}} \right).
$$
Of course, we do not observe $K_t$ and $K_{t-1}$ but noisy proxies $y_t$ and $y_{t-1}$, i.e., the observed case-count on day $t$ and $t-1$ respectively. The hope is that $y_t/y_{t-1}$ is a good proxy for $K_t/K_{t-1}$.  Under SRS with the same sample size at each step, this is true for sufficiently large sample size.  However, similar to Section~\ref{section:rates}, the proxy has error in both numerator and denominator that may cause issues.  We again wish with express the statistical error in the $1+\frac{1}{\gamma} \log \left( \frac{y_t}{y_{t-1}} \right) - R_t$ in useful terms.  Based on Section~\ref{section:rates}, we can re-arrange those terms to show that the error is  given by
$$
\frac{1}{\gamma}\log \bigg[ &1 + \rho_{I_t, K_t} D_{M_t} \sqrt{\frac{1-f}{f}} CV(K_t)  \\
&- \rho_{I_{t-1},K_{t-1}} \sqrt{\frac{1-f}{f}} CV (K_{t-1}) \left[ \rho_{I_t, Y_t} \sqrt{\frac{1-f}{f}} CV(K_{t-1}) + 1 \right] \right].
$$
The coefficient of variation $CV(K_t) = (1- \bar Y_t)$ is the fraction of the uninfected population that remain uninfected.  So here we see the exact same trade-off as with the rates but on the logarithmic scale.  For small values $\log(1+x) = x$ so the biases are similar.  However, the error is no longer scaled by the rate of change in prevalence $\bar Y_2/\bar Y_1$. Instead the error is scaled by the serial interval $\gamma^{-1} = 1/7$.

The above decomposition of the statistical shows us that
